---
title: "Team Assignment 2 All Work"
author: "Ruth Mullin, Hawa Barry, Foqia Shahid"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial setup:
```{r}
library(dplyr)
library(tidyverse)
library(readr)

envsurvey <- read_csv("envsurvey.csv")
```

## Univariate Analysis
```{r}
summary(envsurvey)

# EnvJst
ggplot(envsurvey, aes(y = EnvJst)) +
  geom_boxplot()

# RETH
table(envsurvey$RETH)

# Envlist
table(envsurvey$Envlist)
```

## Questions 1:
```{r}
envsurvey1 <- envsurvey %>%
  mutate(avg_eco = (EnvIss_1+EnvIss_2+EnvIss_6+EnvIss_7+EnvIss_8+EnvIss_13+EnvIss_15+EnvIss_16)/8,
         avg_anthro = (EnvIss_3+EnvIss_4+EnvIss_5+EnvIss_9+EnvIss_10+EnvIss_11+EnvIss_12+EnvIss_14+
                         EnvIss_17+EnvIss_18)/10) %>%
  drop_na(avg_anthro)

# Univariate analysis for avg_eco and avg_anthro
summary(envsurvey1$avg_eco)
summary(envsurvey1$avg_anthro)

ggplot(envsurvey1, aes(y = avg_eco)) +
  geom_boxplot()
ggplot(envsurvey1, aes(y = avg_anthro)) +
  geom_boxplot()

ggplot(data = envsurvey1, aes(x = avg_eco, y = EnvJst)) +
  geom_jitter() +
  geom_smooth()

cor(envsurvey1$avg_eco, envsurvey1$EnvJst)

ggplot(data = envsurvey1, aes(x = avg_anthro, y = EnvJst)) +
  geom_jitter() +
  geom_smooth()

cor(envsurvey1$avg_anthro, envsurvey1$EnvJst) 

envsurvey2 <- envsurvey1 %>%
  mutate(avg_ea = (avg_eco + avg_anthro)/2)

cor(envsurvey2$avg_ea, envsurvey2$EnvJst)

# Average the two averages and check correlation to EnvJst
envsurvey2 <- envsurvey1 %>%
  mutate(avg_ea = (avg_eco + avg_anthro)/2)

cor(envsurvey2$avg_ea, envsurvey2$EnvJst)
```
People who care about ecocentric issues also care more about env justice issues.
People who care about anthro issues might not also care about env justice since the two are not completely related (but anthro = people so we would expect people who are concerned about anthropocentric issues to also care about environmental justice)

## Question 2:
```{r}
# Add RETH to plots above
ggplot(data = envsurvey1, aes(x = avg_eco, y = EnvJst, color = RETH)) +
  geom_jitter() +
  facet_wrap(~ RETH) +
  geom_smooth()

envsurvey1 %>%
  ddply("RETH", summarise, corr=cor(avg_eco, EnvJst))

ggplot(data = envsurvey1, aes(x = avg_anthro, y = EnvJst, color = RETH)) +
  geom_jitter() +
  facet_wrap(~ RETH) +
  geom_smooth()

envsurvey1 %>%
  ddply("RETH", summarise, corr=cor(avg_anthro, EnvJst))
```
Does not really change previous findings (compare how graphs look and correlation coefficients)

## Question 3:
```{r}
# Original graph
ggplot(envsurvey1, aes(x= avg_eco)) +
  geom_density()

# New and improved graph
ggplot(envsurvey1, aes(x = avg_eco, color = Envlist)) +
  geom_density() +
  labs(title = "Environmenalists' Ecocentric Scores are Higher Than Non-environmentalists'", 
       subtitle = "Distribution of Average Ecocentric Scores by Environmental Identity", 
       x = "Average Ecocentric Score", 
       y = "Density", 
       color = "Environmentalist Identity") +
  theme_linedraw()
```

## Question 4:
Relationship between EnvJust, EnvIss, and BYr
```{r}
# Univariate analysis BYr and EnvIss
summary(envsurvey$BYr)
ggplot(envsurvey, aes(x = BYr)) +
  geom_histogram()

summary(envsurvey$EnvIss)
ggplot(envsurvey, aes(y = EnvIss)) +
  geom_boxplot()

# Group BYr by decade
envsurvey <- envsurvey %>%
  mutate(decade = case_when(BYr <= 1949 ~ "1940-1949",
                            BYr <= 1959 & BYr >= 1950 ~ "1950-1959",
                            BYr <= 1969 & BYr >= 1960 ~ "1960-1969",
                            BYr <= 1979 & BYr >= 1970 ~ "1970-1979",
                            BYr <= 1989 & BYr >= 1980 ~ "1980-1989",
                            BYr <= 1999 & BYr >= 1990 ~ "1990-1999"))

# How many people from each decade were surveyed?
table(envsurvey$decade)

# Avg EnvIss by decade
envsurvey %>%
  dplyr::group_by(decade) %>%
  dplyr::summarize(avg_EnvIss = mean(EnvIss))

# Avg EnvJst by decade
envsurvey %>%
  dplyr::group_by(decade) %>%
  dplyr::summarize(avg_EnvJst = mean(EnvJst))

# Plot enviss vs envjst by decade
ggplot(envsurvey, aes(x = EnvIss, y = EnvJst, color = decade)) +
  geom_jitter() +
  facet_wrap(~ decade)

# Corellation coefficient by decade
envsurvey %>%
  dplyr::group_by(decade) %>%
  dplyr::summarize(COR=cor(EnvJst,EnvIss))
```

## Question 5: 
RiskP, avg_eco, Edu
```{r}
# Univariate analysis for Edu
table(envsurvey1$Edu)

ggplot(envsurvey1,aes(x = Edu)) +
  geom_bar() +
  labs(title = "Education Background of Participants",
       subtitle = "Most Participants are well-educated",
       x = "Education",
       y = "Frequency") +
  coord_flip() +
  theme_bw()

# Univariate analysis RiskP 
# To what extent do environmental problems pose a risk to the American public?
ggplot(envsurvey1,aes(y = RiskP)) +
  geom_boxplot() +
  labs(title = "Most participants agree that environmental problems pose a risk to American public",
       subtitle = "Distribution of RiskP score",
       y = "RiskP Score") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#avg_eco: Average for ecocentric questions
ggplot(envsurvey1,aes(y = avg_eco)) +
  geom_boxplot() +
  labs(title = "Most participants recognize environmental issues",
       subtitle = "Distribution of Average Score for ecocentric questions",
       y= "Average Score for Ecocentric Questions") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#Bivariate Analysis
ggplot(envsurvey1, aes(x = avg_eco, y = RiskP)) +
  geom_jitter() +
  geom_smooth() +
  theme_bw() +
  labs(title = "Some correlation between Averge Ecocentric and RiskP Scores",
       subtitle = "r = 0.43 ",
       x = "Average Score for Ecocentric Questions",
       y = "RiskP Score")

cor(envsurvey1$avg_eco, envsurvey2$RiskP)

#Multivariable analysis:
ggplot(envsurvey1, aes(x = avg_eco, y = RiskP, color = Edu)) +
  geom_jitter() +
  theme_bw() +
  labs(title = "Relationship Between Averge Ecocentric and RiskP Scores \n for Different Levels of Education",
       x = "Average Score for Ecocentric Questions",
       y = "RiskP Score") +
  facet_wrap(~Edu) +
  theme(legend.position = "none")

envsurvey1 %>%
  ddply("Edu", summarise, Correlation=cor(avg_eco, RiskP))

 
```

## Conclusion: